---
title: "Case Study - GDP data"
author: "Nimish Sakalle"
date: "October 30, 2016"
output:
  html_document:
    keep_md: yes
  pdf_document: default
  word_document: default
subtitle: Analyzing the GDP of The Top Countries around the world.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = "~/RProject/CaseStudy_fall2016")
knitr::opts_chunk$set(echo = FALSE)
```

<br>

### **CaseStudy - Introduction**

This document is a detail case study of GDP and Education data file from world bank for the top 190 economies of the world.
The data for the case study is downloaded from worldbank website. 
 
  http://data.worldbank.org/data-catalog/GDP-ranking-table
  http://data.worldbank.org/data-catalog/ed-stats

First file GDP.csv has ranking of the countries based on the GDP (Gross Domenstic Product) and second file is detail file with lots of other information about that country (like currency units, groups, debt information etc.)

The objective of this case study is to merge both the files, make the data consistent (including cleanup) and analize data for countries and their GDP group and plot for visual review.

This analysis is done in R (R is a language and environment for statistical computing and graphics).

###**Making environment ready**

``` {r}
## The first step is to load R libraries that are required for the analysis.

library(plyr)
library(ggplot2)
library(knitr)
```

###** Getting Gather, Merge and Tidy file **

```{r}
source("../Analysis/Gather.R", echo=TRUE )
source("../Analysis/Tidy.R", echo=TRUE )
source("../Analysis/Merge.R", echo=TRUE )
```

###**Seting up project directory path.**

**Data Gathering - File read and visual inspection**

``` {r}
setwd("C:\\RProject\\CaseStudy_fall2016\\Data")
# File Reading
GDP <- read.csv("GDP.csv", skip=3,header=TRUE)
# Checking for file header names.
  names(GDP)
#Its important to save original file layout and data into vector, this may come handy to have original format. 
  GDP_RAW<-GDP
##Below commands to  read second file.
  FEDSTATS <- read.csv("EDSTATS.csv",header=TRUE)
# Like the first file, below steps to check the of second file.
# This is to get the number of rows in file.
  NROW(FEDSTATS)
#Its important to save original file layout and data into vector, this may come handy to have original format.
  FEDSTATS_RAW<-FEDSTATS
```

###**Tidy Data**

 Data quality is most important aspect of any sort of data analysis.
 Below are steps to make sure data is cleaned up for good anaylsis. Its important to review data before and after data cleanup.
 
``` {r}
### Step3: file and data quality checks and cleanup.
##  First and fifth column does not have any name. below command it to add the column header.

  colnames(GDP)[1] <- "CountryCode"
  colnames(GDP)[5] <- "GDP"

# As check above columms x.1,x.2,x.3,x.4,x.5,x.6 does not have any data, these columns can be dropped from the dataset.
  GDP.Final <- GDP[,c(1:2,4:5)]
  NROW(GDP.Final)

# Cleaning up of GDP File. Ranking attributes does have null values and below command will eliminate null data frame.
  GDP.Final[GDP.Final == ""] <- NA
  GDP.Final.Clean <- GDP.Final[!(is.na(GDP.Final$Ranking)), ]

# removing rows where the column values are null from attribute GDP.

  Final.GDP <- GDP.Final.Clean[!(is.na(GDP.Final.Clean$GDP)), ]
  nrow(Final.GDP)
```
The above show the numbers of rows in the data frame before and after cleanup.

###**Data Merge and Analysis 1**

*Data Merge using country code from both the files.*
*verify Numbers of Ids matched from both the files.*

To continue with the analysis, in the following steps data from GDP and Education file is merged on country code.
After merging both the file, its important to visually inspect data to make sure data does not have any more anomalies.

``` {r}
# This is to merge both the file using "COuntryCode" COlumn

  Merge.Data <- merge(Final.GDP, FEDSTATS, BY = "CountryCode")
  
# below command is to copy only the required columns to the dataframe. this is useful to eliminate colums not required for analysis.

  Merge.GDP.STAT <- Merge.Data[,c(1,2,3,4,6)]
  names(Merge.GDP.STAT)
# Below command output shows the number of IDs matched.
  
  dim(Merge.GDP.STAT)
  GDP.STAT.MATCH <- nrow(Merge.GDP.STAT)
  print(paste0("IDs Matched:", GDP.STAT.MATCH))
```

#####Number of ID matched 189


###**Data Sort and Analysis 2**

To sort the data on GDP in assending order and publish the coutry stand 13 on the list.
```{r}
# To sort the data by GDP, this column has to convert into numeric. This column is currently defined as character.
  
  Merge.GDP.STAT$GDP <-as.numeric(gsub("[^[:digit:]]","", Merge.GDP.STAT$GDP))
  Merge.GDP.STAT$Ranking <-as.numeric(gsub("[^[:digit:]]","", Merge.GDP.STAT$Ranking))
 
# Below command is to sort the data by GDP.   
  Sort.GDP.Stat <- Merge.GDP.STAT[order(Merge.GDP.STAT$GDP), ]

# This is to dispaly 13 row of the data set\ after sorting the dataset.  
  Sort.GDP.Stat.13 <- Sort.GDP.Stat[13,]
  knitr::kable(Sort.GDP.Stat.13)
```


###**Analysis 3: AVG GDP Ranking **

Getting a mean of ranking based on the income group.

```{r}
#  The below command is to provide the mean ranking by income group. 
  Merge.GDP.STAT.AVG  <- ddply(Merge.GDP.STAT, .(Income.Group), summarize,  Ranking=mean(Ranking))
# Below data set is a mean ranking by income group.
  knitr::kable(Merge.GDP.STAT.AVG)
```

The mean Ranking of high Income OECD and Non OECD group is:

|Income.Group         |   Ranking|
|:--------------------|---------:|
|High income: nonOECD |  91.91304|
|High income: OECD    |  32.96667|

###**Analysis 4** 

To generate a plot to see GDP on scale based on the income group.

```{r}
## ggplot with color by income group.
## This below command it to create QQ plot.

ggplot(Merge.GDP.STAT)+ geom_point(aes(x=GDP,y=Income.Group,colour=Income.Group))+scale_x_log10()+labs(x = Merge.GDP.STAT$GDP, y= Merge.GDP.STAT$Income.Group)

```


This plot shows GDP number (Log-scale 10) of the countries based on the income group. This graphs clearly show the mean GDP of income group with  Highincome OECD and High Income Non OECD is significantly higher than other income groups.

###**Analysis 5**

To cut the countries into 5 quantil for each income group and see the data distribution.

```{r}
# Below command cut GDP into 5 quantile
Merge.GDP.STAT$Quantile<- cut(Merge.GDP.STAT$Ranking,breaks=5) 
#take the quantiles and income.group from the merge data file and create a table 
quant.table<-table(Merge.GDP.STAT$Income.Group, Merge.GDP.STAT$Quantile) 
knitr::kable(quant.table)
```


The above are number of countiries belongs ito each income group divided into five quantiles. This data clearly show GPD of the counties is distributed, even that are part of any particular income group.  

```{r}
LMI.Highgdp<- Merge.GDP.STAT[which(Merge.GDP.STAT$Ranking <= 38 & Merge.GDP.STAT$Income.Group == "Lower middle income"),]
```

**List of lower income countries in top 38 nations.**

```{r}
# List of lower income countries in top 38 nations.

knitr::kable(LMI.Highgdp)
```

*The above 5 Lower income countries belowgs amongst 38 nations with highest GDP.

### **CaseStudy - Conclusion**

This objective of this case study is to perform key steps that requires in any data analysis.
These keys steps are very foundation of any project where it requires to read, study, interpret and process information.
We performed foundation step like data gathering, data cleaning, data merging, data analysis and data processing on data GDP and Education data from world bank.
One more very important step (making reproducable research and data) is been accompleshed inthis case study, by making data and research available at GitHub.
All the above steps are performend in R using R studio.

By looking at the data and graphs it is very clear that GDP of some of the courtries are different than their income group.
SOme of the contries belong to low income group have high GDP and vice versa.



